# Simulation

All previous chapters were concerned with defining a compartmental model. In this chapter we switch to getting results from a defined model.

Once a model object is defined, it can be used to generate simulations using the `simulation_history` function.

```{r eval=FALSE}
simulation_history(sir)
```

```{r echo = FALSE}
rmarkdown::paged_table(simulation_history(sir))
```

The output contains a column for the simulation date, one column for each state variable (`S`, `I` and `R` in this case), and one column for every time-varying rate (`S_to_I`). The names of the time-varying rates are always of the form `{from_state}_to_{to_state}`. The reason why `S_to_I` is time-varying in this model is that it depends on a state variable, `I`, which is itself varying at every time-step. The `rate_summary` function can be used to remind us of this fact.

```{r}
(sir
  %>% rate_summary(include_formula = TRUE) 
  %>% select(from, to, formula)
)
```

We see here that `S_to_I` does indeed depend on `I` in its formula, whereas `I_to_R` depends only on a parameter, `gamma`.

Note that the above command uses the tidyverse-style pipe, `%>%`, operator and another tidyverse function, `select`. This illustrates a general philosophy of McMasterPandemic, which is that we try to make the outputs plug into other existing and popular tools rather than reinvent existing functionality for a narrower purpose. For example, the `rate_summary` function returns a data frame that can be manipulated by other data frame manipulation tools.

We can plug into other existing and popular tools to make a plot of the simulated epidemic trajectory.

```{r}
(sir
 %>% simulation_history
 %>% select(-S_to_I)
 %>% pivot_longer(-Date, names_to = "State", values_to = "Population")
 %>% mutate(State = factor(State, levels = topological_sort(sir)))
 %>% ggplot
 + geom_line(aes(Date, Population, colour = State))
)
```

::: {.infobox .caution data-latex="{caution}"}
There are a few places you can go from here:

1. Learn how to fit a model to observed data through [Calibration]
2. Learn how to modify the values of parameters in simulation time using [Time Varying Parameters]
3. Keep reading to learn about simulating with [Observation Error]
:::

## Observation Error

```{r obs_err_sims}
sir_with_obs_err = (sir
  %>% update_loss_params(data.frame(
    Parameter = "nb_disp", 
    Distribution = "negative_binomial",
    Variable = c("S", "I", "R")
  ))
  %>% update_params(c(
    nb_disp_S = 1e4, 
    nb_disp_I = 1e4, 
    nb_disp_R = 1e4
  ))
)
```

```{r plot_obs_err}
set.seed(1L)
(sir_with_obs_err
  %>% simulation_history(obs_error = TRUE)
  %>% select(Date, S, I, R)
  %>% pivot_longer(-Date, names_to = "var", values_to = "value")
  %>% rename(date = Date)
  %>% mutate(var = factor(var, levels = topological_sort(sir)))
  %>% ggplot
  + facet_wrap(~ var)
  + geom_point(aes(date, value))
)
```
