# Simulation

```{r echo = FALSE}
sir = (flexmodel(
  params = c(beta = 0.1, gamma = 0.01, N = 100),
  state = c(S = 99, I = 1, R = 0),
  start_date = "2020-03-11",
  end_date = "2020-12-01",
  do_make_state = FALSE
)
  %>% add_rate("S", "I", ~ (I) * (beta) * (1/N))
  %>% add_rate("I", "R", ~ (gamma))
)
```

All previous chapters were concerned with defining a compartmental model. In this chapter we switch to getting results from a defined model.

Once a model object is defined, it can be used to generate simulations using the `simulation_history` function.

```{r eval=FALSE}
simulation_history(sir)
```

```{r echo = FALSE}
rmarkdown::paged_table(simulation_history(sir))
```

The output contains a column for the simulation date, one column for each state variable (`S`, `I` and `R` in this case), and one column for every time-varying rate (`S_to_I`). The names of the time-varying rates are always of the form `{from_state}_to_{to_state}`. The reason why `S_to_I` is time-varying in this model is that it depends on a state variable, `I`, which is itself varying at every time-step. The `rate_summary` function can be used to remind us of this fact.

```{r}
(sir
  %>% rate_summary(include_formula = TRUE) 
  %>% select(from, to, formula)
)
```

We see here that `S_to_I` does indeed depend on `I` in its formula, whereas `I_to_R` depends only on a parameter, `gamma`.

Note that the above command uses the tidyverse-style pipe, `%>%`, operator and another tidyverse function, `select`. This illustrates a general philosophy of McMasterPandemic, which is that we try to make the outputs plug into other existing and popular tools rather than reinvent existing functionality for a narrower purpose. For example, the `rate_summary` function returns a data frame that can be manipulated by other data frame manipulation tools.

We can plug into other existing and popular tools to make a plot of the simulated epidemic trajectory.

```{r}
(sir
 %>% simulation_history
 %>% select(-S_to_I)
 %>% pivot_longer(-Date, names_to = "State", values_to = "Population")
 %>% mutate(State = factor(State, levels = c("S", "I", "R")))
 %>% ggplot
 + geom_line(aes(Date, Population, colour = State))
)
```
