# Other Variables

So far the only types of variables that we have discussed are parameters and state variables. In this chapter we cover all of the other kinds of variables.

There are two basic kinds of other variables: (1) [Intermediate Results] and (2) [Additional Variables in the Simulation History]. The main difference between the two types is that the former can be used for more purposes than the latter. In particular, intermediate results can be used in expressions that define rate matrices whereas additional variables in the simulation history cannot. Conversely, all intermediate results get added to the simulation history.

## Intermediate Results

Sometimes the expression giving the rate of flow between two compartments is quite complex, or even just impossible to express in terms of just parameters and state variables given the restrictions described in the chapter on [Flows Between States]. In these cases, it is often convenient or even necessary to store intermediate combinations of parameters and state variables and then use these combinations in subsequent expressions for flow rates.  There are two basic kinds of intermediate results: (1) sums of state variables and parameters and (2) more general expressions, called `factr`s, that combine state variables, parameters, and their sums.

### Sums of State Variables and Parameters

One may save any sum of state variables and/or parameters to the model. In our SIR model for example, we specified the total population size in the parameter vector as `N`. However, we could compute this population size using the `add_state_param_sum` function.
```{r add_sum}
sir = (flexmodel(
  params = c(beta = 0.1, gamma = 0.01),
  state = c(S = 99, I = 1, R = 0),
  start_date = "2020-03-11",
  end_date = "2020-12-01"
)
  %>% add_state_param_sum("N", "S|I|R")
  %>% add_rate("S", "I", ~ (I) * (beta) * (1/N))
  %>% add_rate("I", "R", ~ (gamma))
)
```

The first argument to the `add_state_param_sum` is just a name for the sum. Note that this name is used in the `add_rate` function that defines the flows from `S` to `I`. The second argument to the `add_state_param_sum` is a regular expression that is matched agains all parameter names and state variable names. Here we match any state variable with `"S|I|R"`, which tells McMasterPandemic to add up all state variables and save the result in the variable called `"N"`.


### Factrs

First off, I'm sorry about the name `factr`. It is not a typo, I just don't have the energy to change it to something better right now.

`factr`s are simply a saved expression that follows the same rules for rates of [Flows Between States]. These `factr`s are named and can subsequently be used in rate expressions.

The SIR model is not really complex enough to justify using `factr`s, but if we wanted to we could store the proportion of individuals that are in the `I` box as a `factr` and then use that `factr` in the expression for the rate from `S` to `I`.
```{r add_factr}
sir = (flexmodel(
  params = c(beta = 0.1, gamma = 0.01),
  state = c(S = 99, I = 1, R = 0),
  start_date = "2020-03-11",
  end_date = "2020-12-01"
)
  %>% add_state_param_sum("N", "S|I|R")
  %>% add_factr("Iprop", ~ (I) * (1/N))
  %>% add_rate("S", "I", ~ (Iprop) * (beta))
  %>% add_rate("I", "R", ~ (gamma))
)
```

## Additional Variables in the Simulation History

The `simulation_history` function returns a data frame containing many of the variables in the model, with each row giving each time step in the simulation. This `simulation_history` function is introduced in the chapter on [Simulation]. This simulation history table is used for two purposes: (1) to return to the user as simulation output and (2) to compare with observed time series in in the process of [Calibration].

The simulation history contains the following variables in the following order:

1. State variables
2. Flow rates that vary in simulation time (constant rates are omitted)
3. Sums of state variables and parameters
4. `factr`s
5. Expressions following the rules in [Flows Between States] of any of the variables in 1-4 above
6. Lagged differenced versions of 1-5
7. Convolutions of 1-5

Items 5-7 consitute the additional variables in the simulation history. In the next sections we illustrate how to add such variables to the SIR model.

### Simulation History Expressions

TODO

### Lagged Differencing

TODO

### Convolutions

TODO
