# Time Varying Parameters

Any parameter in a `McMasterPandemic` model can be scheduled to vary in time. A time-variation schedule is a data frame with one row for each date on which each parameter changes its value. This data frame has four columns:

* `Date` -- Date on which a particular parameter changes its value
* `Symbol` -- String giving the symbol representing the changing parameter
* `Value` -- The numeric value used to change the value of the parameter, the effect of which depends on the value of the `Type` column
* `Type` -- One of the following strings:
  * `"abs"` -- The `Value` column is the new value for the parameter on `Date`
  * `"rel_orig"` -- The `Value` column is multiplied by the original value of the changing parameter at the beginning of the simulation, to generate a new value for the parameter on `Date`
  * `"rel_prev"` -- The `Value` column is multiplied by the previous value of the changing parameter, to generate a new value for the parameter on `Date`

Once a time-variation data frame is produced it can get added to a model when it is created via the `params_timevar` argument of `flexmodel`, or by updating an existing `flexmodel` object using the `update_piece_wise` function.

## Example of Time-Variation

In this example we create a time-variation schedule that causes the transmission, `beta`, drop to very low levels on May 15.
```{r random_timevar}
random_timevar = data.frame(
  Date = ymd(20200515),
  Symbol = 'beta',
  Value = 0.01,
  Type = 'abs'
)
random_timevar
```

```{r add_timevar}
sir_with_timevar = (sir
  %>% update_piece_wise(random_timevar)
)
```


```{r simulate_with_timevar}
timevar_sims = (sir_with_timevar
  %>% simulation_history(include_initial_date = FALSE)
  %>% tidyr::pivot_longer(-Date, names_to = "var")
  %>% rename(date = Date)
  %>% mutate(value = round(value))
  %>% filter(var %in% c("S", "I", "R"))
)
(ggplot(timevar_sims)
  + geom_line(aes(date, value, colour = var))
)
```

Notice the abrupt break-point on May 15.

See [The SIRV model] and [Covid SEIR] for other examples of the use of time-varying parameters in simulation models.

## Calibrating Time-Variation Schedules

Entries in the `Value` column can be `NA`, indicating that these should be fitted using [Calibration]. When flagging time-variation values for calibration in this way, one must also provide information on any parameter transformations and prior distributions.  Do provide this information, we use a technique similar to the one described in the [Calibration] chapter to specify transformations and prior distributions for parameters in the `params` element of `flexmodel` objects. In that chapter we used the `add_opt_params` function -- here we use the `add_opt_tv_params` function (the `tv` stands for time-variation).

To illustrate calibration of time-variation values, we mark for calibration the break-point on May 15 in our example above.

```{r set_timevar_na}
calibrate_timevar = (random_timevar
  %>% mutate(Value = NA)
)
```

Then we update our model with the `sir_with_timevar` simulations to fit to, and specify how to optimize the parameters.
```{r cal_tv}
sir_to_cal_tv = (sir_with_timevar
  %>% update_observed(timevar_sims)
  %>% update_piece_wise(calibrate_timevar)
  %>% add_opt_params(log_beta ~ log_flat(-1)
    #, log_nb_disp_S ~ log_normal(0, 5)
    #, log_nb_disp_I ~ log_normal(0, 5)
    #, log_nb_disp_R ~ log_normal(0, 5)
  )
  %>% add_opt_tv_params(tv_type = "abs"
    , log_beta ~ log_flat(-1)
  )
)
```

The key function here is `add_opt_tv_params`, which allows us to specify a flat prior on the log scale for fitting the time-variation values.

It turns out that in this case we need to make one technical adjustment -- see [Simulated time-series close to zero] for an explanation.

```{r do_sim_constraint}
sir_to_cal_tv$do_sim_constraint = TRUE
```

With this `sir_to_cal_tv` object we can now fit this model to the simulated data.

```{r}
sir_cal_tv = calibrate_flexmodel(sir_to_cal_tv, optimizer = 'nlminb')
```

The time-variation values before, during, and after optimization look as we would expect, given that we fitted to the same model that generated the data.
```{r}
c(
  before = sir_with_timevar$timevar$piece_wise$schedule$Value,
  during = sir_to_cal_tv$timevar$piece_wise$schedule$Value,
  after = sir_cal_tv$timevar$piece_wise$schedule$Value
)
```

## Scenario Exploration -- Forecasting with Time-Variation Schedules

## Complex Time-Variation Schedules
