# Ensemble Forecasts

```{r redo_obs_err_fit, echo = FALSE, eval = FALSE}
sir_with_obs_err = (flexmodel(
    params = c(beta = 0.1, gamma = 0.01, N = 100),
    state = c(S = 99, I = 1, R = 0),
    start_date = "2020-03-11",
    end_date = "2020-12-01",
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (I) * (beta) * (1/N))
  %>% add_rate("I", "R", ~ (gamma))
  %>% update_loss_params(data.frame(
    Parameter = "nb_disp", 
    Distribution = "negative_binomial",
    Variable = c("S", "I", "R")
  ))
  %>% update_params(c(
    nb_disp_S = 1e4, 
    nb_disp_I = 1e4, 
    nb_disp_R = 1e4
  ))
)
set.seed(1L)
noisy_data = (sir_with_obs_err
  %>% simulation_history(include_initial_date = FALSE, obs_error = TRUE)
  %>% select(Date, I)
  %>% pivot_longer(-Date, names_to = "var", values_to = "value")
  %>% rename(date = Date)
)
sir_to_calibrate = (sir_with_obs_err
  %>% update_observed(
    noisy_data, 
    loss_params = loss_params(sir_with_obs_err)
  )
  %>% update_opt_params(
    log_beta ~ log_flat(0),
    log_nb_disp_S ~ log_normal(10, 0.01), # fix at ~Poisson error
    log_nb_disp_I ~ log_normal(0, 5),     # free parameter
    log_nb_disp_R ~ log_normal(10, 0.01)  # fix at ~Poisson error
  )
)
sir_calibrated = calibrate_flexmodel(sir_to_calibrate)
```

Good forecasts include measures of uncertainty associated with each forecasted data point. The `simulate_ensemble` function allows one to make such forecasts. In [Calibrating with Observation Error] we fitted a model to simulated data with observation error. Here we use this model to illustrate the use of the `simulate_ensemble` function.

The first thing to do before making a forecast is to extend the end date, so that we are actually forecasting the future.
```{r extend_end_date}
sir_obs_err_to_forecast = extend_end_date(
  sir_obs_err_calibrated, 
  days_to_extend = 100
)
```

Next just pass the model that is ready for forecasting to the `simulate_ensemble` function.
```{r simulate_ensemble, eval = TRUE}
obs_err_forecasts = (simulate_ensemble(
    sir_obs_err_to_forecast, 
    use_progress_bar = FALSE
  )
  %>% filter(var != "S_to_I")
)
```
```{r paged_obs_err_forecasts, echo = FALSE}
rmarkdown::paged_table(obs_err_forecasts)
```

The forecasts can be joined back to the observed data to make plots (TODO: fix the naming of the columns of these functions to be more consistent).
```{r join_obs_data_to_sims, eval = TRUE}
obs_err_fits = (obs_err_forecasts
  %>% left_join(
    noisy_data, 
    c("Date" = "date", "var" = "var"), 
    suffix = c("_fitted", "")
  )
  %>% mutate(
    var = factor(
      var, 
      levels = topological_sort(sir_obs_err_to_forecast))
  )
)
(ggplot(obs_err_fits)
  + facet_wrap(~var, scales = 'free', nrow = 2)
  + geom_ribbon(aes(x = Date, ymax = upr, ymin = lwr), alpha = 0.5)
  + geom_point(aes(Date, value), alpha = 0.7)
  + geom_line(aes(Date, value_fitted), colour = "red")
)
```


## Time-Varying Ensemble Forecasts
