# TMB Engine

The initial engine for calibrating and forecasting in McMasterPandemic was R itself. In the [refactored](#fast-and-flexible-modelling-with-mcmasterpandemic) version of McMasterPandemic we use template model builder, or [TMB](https://github.com/kaskr/adcomp). It is possible to interact with TMB objects directly using the `McMasterPandemic::tmb_fun` function, which we illustrate in this section to do MCMC simulation.

```{r echo = FALSE, eval = FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]        
  S [label = 'S']
  I [label = 'I']
  
  # edge definitions with the node IDs
  S -> I
  }",
  height = 100, width = 200)
```


HACK: For some reason `tmbstan` is not working unless we compile the C++ code rather than use the package-compiled objects.

```{r system_file_cpp}
cpp_dir = system.file('tmb', '0.2.0', package = "McMasterPandemic")
set_spec_version('0.2.0', cpp_dir, use_version_directories = FALSE)
```

We can get the TMB object by calling the `tmb_fun` function on a `flexmodel_to_calibrate` object.
```{r get_tmb_object}
sir_obs_err_tmb = tmb_fun(sir_obs_err_to_calibrate)
```

We can pass this `tmb_fun` object to `tmbstan` (in the [tmbstan](https://github.com/kaskr/tmbstan) package) to generate MCMC samples using [rstan](https://mc-stan.org/users/interfaces/rstan).
```{r mcmc}
sir_obs_err_stan = tmbstan(
  sir_obs_err_tmb,
  chains = 1
)
names(sir_obs_err_stan) = c(
  names(tmb_params_trans(sir_obs_err_to_calibrate)), 
  "log_posterior"
)
traceplot(sir_obs_err_stan)
sir_obs_err_stan
```

One may also use [shinystan](https://mc-stan.org/users/interfaces/shinystan) with these objects.
