# Examples

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hello World: Simulating an SIR Model

```{r hello_world, fig.height=8, fig.width=8}
state = c(S = 20000, I = 100, R = 0)
sir_model = (
  flexmodel(
    params = c(
      gamma = 0.06,
      beta = 0.15,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = FALSE
  )
  %>% add_rate("S", "I", ~ (1/N) * (beta) * (I))
  %>% add_rate("I", "R", ~ (gamma))
)
sir_model
```

```{r echo=FALSE,eval=FALSE}
library(visNetwork)

state_nms = names(sir_model$state)
nn = visNetwork(
  nodes = data.frame(
    id = state_nms,
    label = state_nms,
    title = state_nms,
    physics = FALSE),
  edges = select(rate_summary(sir_model), from, to) %>% mutate(arrows = "to")
) %>%
  visLegend() %>%
  visPhysics(stabilization = FALSE) %>% 
  visInteraction(
    dragNodes = FALSE,
    dragView = FALSE,
    zoomView = FALSE
  )
nn
```

```{r print_sir}
(sir_model
    %>% simulation_history
    %>% select(-S_to_I)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% mutate(epi_cat = factor(epi_cat, levels = topological_sort(sir_model)))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

```{r plot_sir}
(sir_model
    %>% simulation_history
    %>% rename(`force of infection` = S_to_I)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`))
  )
```

One may use this functionality to compute basic epidemiological parameters such as $R_0$, $r$, and $\bar{G}$. We do this by modifying the model slightly, simulating, and summarizing the output.  TODO: actually compute these parameters.

```{r epi_pars, echo=TRUE, eval=FALSE}
epi_pars(sir_model
  , exposed_state_nm = "I"
  , foi_nm = "S_to_I"
  , N = 1
)
```

```{r R0_r_Gbar, eval = FALSE, echo = FALSE}
sir_model_for_R0 = sir_model

# start from a single infected individual
state_default(sir_model_for_R0) = c(S = 0, I = 1)
pars_base_sim(sir_model_for_R0) = c(N = 1)

sir_epi_pars = (sir_model_for_R0
  
  # run simulation to convergence
  %>% extend_end_date(300)
  %>% simulation_history
  
  %>% summarise(
    
    # reproductive number
    R0 = sum(S_to_I),
    
    # mean generation interval
    Gbar = sum(seq_along(S_to_I) * S_to_I / R0),
    
    # intrinsic rate of increase
    r = uniroot(make_lotka_euler(S_to_I), c(0, 100))$root
  )
)

# R0 by standard SIR model theory
with(as.list(pars_base_sim(sir_model)), beta/gamma)

# R0 and other stats by kernel theory
sir_epi_pars

(sir_model
  %>% simulation_history
  %>% select(Date, S_to_I, S)
  %>% mutate(Rt = sir_epi_pars$R0 * (S / pars_base_sim(sir_model)[["N"]]))
  %>% ggplot
  + geom_line(aes(Date, Rt))
)
```

## SI

```{r si}
si = (flexmodel(
    params = c(beta = 0.15, gamma = 0.06, N = 100),
    state = c(S = 99, I = 1),
    start_date = "2000-01-01", end_date = "2000-06-01"
  )
  %>% add_rate("S", "I", ~ (I) * (beta) * (1/N))
  %>% add_rate("I", "S", ~ (gamma))
  %>% add_factr("ratio", ~ (gamma) * (1/beta))
  %>% add_factr("S_hat", ~ (N) * (ratio))
  %>% add_factr("I_hat", ~ (N) * (1 - ratio))
)
(si
  %>% simulation_history
  %>% select(-S_to_I, -ratio)
  %>% pivot_longer(-Date, names_to = "State", values_to = "Population")
  %>% separate(State, c("State", "Equilibrium"), "_")
  %>% mutate(Equilibrium = ifelse(is.na(Equilibrium), 'no', 'yes'))
  %>% ggplot
  + geom_line(aes(Date, Population, colour = State, linetype = Equilibrium))
)
```

## SEIR

```{r seir}
state = c(S = 20000, E = 0, I = 100, R = 0)
seir_model = (
  flexmodel(
    params = c(
      alpha = 0.05,
      gamma = 0.06,
      beta = 0.15,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = TRUE,
  )
  %>% add_rate("S", "E", ~ (1/N) * (beta) * (I))
  %>% add_rate("E", "I", ~ (alpha))
  %>% add_rate("I", "R", ~ (gamma))
)
seir_model
```

## Structure: Two-Strain SIR

```{r two_strain_sir}

strains = c("wild", "variant")
state = c(
  S = 20000, 
  I_wild = 49, I_variant = 1,
  R_wild = 0,   R_variant = 0
)
two_strain_model = (
  flexmodel(
    params = c(
      gamma = 0.06,       
      beta_wild = 0.15,   
      beta_variant = 0.25,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = TRUE
  )
  %>% vec_rate(
    "S",
    "I" %_% strains, 
     vec("beta" %_% strains) * struc("1/N") * vec("I" %_% strains)
  )
  %>% rep_rate("I", "R", ~ (gamma))
)
```

```{r plot_two_strain}
regex = "^(S|I|R)(_*)(|wild|variant)$"
(two_strain_model
    %>% simulation_history
    %>% select(Date, matches(regex))
    %>% pivot_longer(!Date)
    %>% rename(state = value)
    %>% mutate(strain  = sub(pattern = regex, replacement = "\\3", name))
    %>% mutate(epi_cat = sub(pattern = regex, replacement = "\\1", name))
    %>% mutate(strain = factor(strain, c("", "wild", "variant")))
    %>% mutate(epi_cat = factor(epi_cat, c("S", "I", "R")))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat, linetype = strain))
  )
```

```{r plot_two_strain_again}
(two_strain_model
    %>% simulation_history
    %>% pivot_longer(starts_with("S_to_I"))
    %>% mutate(name = sub("S_to_I_", "", name))
    %>% rename(`force of infection` = value)
    %>% rename(strain = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`, colour = strain))
  )
```

## Erlang SEIR

David, Jonathan, and David describe the [Erlang SEIR model](https://epubs.siam.org/doi/pdf/10.1137/18M1186411) in continuous time. Here is a discrete time version of it.
```{r erlang}
n = 4 # number of I states
m = 6 # number of E states
erlang_seir = (flexmodel(
  # FIXME: only working for no demography (so mu = 0 for now)
  params = c(mu = 0, beta = 1.5, m = m, n = n,
             gamma = 1.2, sigma = 0.1),
  state = c(
    S = 1-1e-3,
    layered_zero_state("E" %_% 1:m),
    I_1 = 1e-3,
    layered_zero_state("I" %_% 2:n),
    R = 0,
    D = 0,
    . = 1),
  start_date = "2000-01-01",
  end_date = "2000-04-01",
  do_hazard = TRUE
)
  %>% add_state_param_sum("I", "^I_[0-9]+")

  # birth
  %>% add_rate(".", "S", ~ (mu))

  # death
  %>% rep_rate(
    from = c("S", "E" %_% 1:m, "I" %_% 1:n, "R"), 
    to = "D",
    formula = ~ (mu))
  
  # infection
  %>% add_rate("S", "E_1", ~ (beta) * (I))

  # become infectious
  %>% add_rate("E" %_% m, "I_1", ~ (m) * (sigma))

  # sojourn through exposed compartments
  %>% rep_rate(
    "E" %_% 1:(m-1),
    "E" %_% 2:m,
    ~ (m) * (sigma)
  )

  # sojourn through infectious compartments
  %>% rep_rate(
    "I" %_% 1:(n-1),
    "I" %_% 2:n,
    ~ (n) * (gamma)
  )

  # recovery
  %>% add_rate("I" %_% n, "R", ~ (n) * (gamma))

  # nothing flows out of . because it is a dummy
  # state used to generate flows that are not per-capita
  %>% add_outflow("[^.]")
)

(erlang_seir
  %>% simulation_history
  %>% select(-., -D, -S_to_E_1)
  %>% pivot_longer(-Date, names_to = "State", values_to = "Count")
  %>% mutate(State = sub("_[0-9]+", "", State))
  %>% group_by(Date, State)
  %>% summarise(Count = sum(Count))
  %>% ungroup
  %>% mutate(State = factor(State, c("S", "E", "I", "R")))
  %>% ggplot() + geom_line(aes(Date, Count, colour = State))
)

# check to make sure that the population density
# remains constant at one
(erlang_seir
  %>% simulation_history
  %>% select(-Date, -., -D, -S_to_E_1, -I)
  %>% rowSums
  %>% sapply(all.equal, 1L)
  %>% sapply(isTRUE)
  %>% all
)
```

## SIRV

```{r sirv}
state = c(S = 20000, I = 100, R = 0, V = 0)
params = c(
      gamma = 0.06,
      beta = 0.15,
      v = 0, # initial vaccination rate
      N = sum(state)
    )

# roll out the vaccine by bumping the 
# vaccination rate twice
params_timevar = data.frame(
  Date = c("2020-02-01", "2020-03-01"), 
  Symbol = c("v", "v"),  
  Value = c(0.01, 0.1), 
  Type = c("abs", "abs"))

sirv_model = (
  flexmodel(
    params = params,
    state = state,
    params_timevar = params_timevar,
    start_date = "2020-01-01",
    end_date = "2020-05-01",
    do_hazard = TRUE
  )
  %>% add_rate("S", "I", ~ (beta) * (1/N) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_rate("S", "V", ~  (v))
)

(sirv_model
    %>% simulation_history
    %>% select(Date, matches("^(S|I|R|V)$"))
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% mutate(epi_cat = factor(epi_cat, topological_sort(sirv_model)))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
     +  geom_vline(aes(xintercept = as.Date(Date)), data = params_timevar, colour = 'lightgrey')
  )

```


## Variolation model

The [variolation model](https://royalsocietypublishing.org/doi/10.1098/rsif.2021.0781) is ... TODO: add description.

```{r variolation}
state = c(
  S = 20000, 
  I_severe = 50, I_mild = 50,
  R_mild = 0,   R_severe = 0
)
params = c(
  nu = 0.0105, # birth rate
  mu = 0.0105, # death rate
  delta = 1/7, # waning immunity rate
  gamma_mild = 1/(3.3 + 9),  # recovery rate of mild cases
  gamma_severe = 1/(3.3 + 14), # recovery rate of severe cases
  beta_mild = 0.15,  # transmission rate of infections leading to mild cases
  beta_severe = 0.3, # transmission rate of infections leading to severe cases
  m = 0.6 # probability of developing mild illness
)

# model structure
all_states = names(state)
severity = c("mild", "severe")
beta_vec = vec("beta" %_% severity)
I_vec = vec("I" %_% severity)
foi = kronecker(vec("m", "1-m"), sum(beta_vec * I_vec * struc("1/N")))

variolation_model <- (
  flexmodel(
    params = params,
    state = state,
    start_date = "2020-01-01",
    end_date = "2020-04-01"
  )
  %>% add_state_param_sum("N", any_var(state))
  
  # births and deaths 
  # (FIXME: this only works if nu = mu)
  %>% rep_rate(
    all_states,
    "S",
    ~ (nu)
  )
  
  # infection
  %>% vec_rate(
    "S",
    "I" %_% severity, 
     foi
  )
  
  # waning immunity
  %>% rep_rate(
    "R" %_% severity,
    "S",
    ~ (delta)
  )
  
  # recovery
  %>% vec_rate(
    "I" %_% severity,
    "R" %_% severity,
    vec("gamma" %_% severity)
  )
)

(variolation_model
    %>% simulation_history
    %>% select(Date, matches(any_var(variolation_model$state)))
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )

```

## SEIRD

This is the [Mac Theo Bio Model](https://github.com/mac-theobio/mac-theobio.github.io/blob/master/seird.jpg).

```{r seird}
state = c(S = 20000, E = 50, I = 50, R = 0, D = 0, Null = 0)
params = c(
  # transmission rates of susceptibles from live and dead individuals
  beta_I = 5,
  beta_D = 2,
  
  # average times in various boxes
  T_E = 10,
  T_I = 14,
  T_D = 10,
  
  # probability of death given infection
  f = 0.5
)
seird_model = (
  flexmodel(
    params = params, 
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-03-01",
    do_hazard = TRUE
  )
  %>% add_state_param_sum("N", any_var(state))
  %>% add_rate("S", "E", ~ (beta_I) * (1/N) * (I) + (beta_D) * (1/N) * (D))
  %>% add_rate("E", "I", ~ (1/T_E))
  %>% add_rate("I", "R", ~ (1/T_I) * (1 - f))
  %>% add_rate("I", "D", ~ (1/T_I) * (f))
  %>% add_rate("D", "Null", ~ (1/T_D))
)
(seird_model
    %>% simulation_history
    %>% select(Date, matches(any_var(state)))
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% mutate(epi_cat = factor(epi_cat, levels = topological_sort(seird_model)))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

## Covid SEIR

The BC covid modelling group uses this [compartmental model](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008274) for their inference and forecasting work.  This model can be expressed in McMasterPandemic with the following.

```{r covid_seir}
foi =  (
    vec(c("1", "f")) 
  * struc("beta") 
  * struc("1/N") 
  * (struc("(I) + (E2)") + (struc("f") * struc("(Id) + (E2d)")))
)
params = c(
  N = 5100000,
  D = 5,
  k1 = 0.2,
  k2 = 1,
  q = 0.05,
  ud = 0.1,
  ur = 0.02,
  psir = 0.3,
  shape = 1.73,
  scale = 9.85,
  beta = 0.433,
  f = 1
)
state = c(
  S = 849999,
  E1 = 0.53,
  E2 = 0.13,
  I = 0.67,
  Q = 0,
  R = 0,
  Sd = 4249993,
  E1d = 2.67,
  E2d = 0.67,
  Id = 3.33,
  Qd = 0,
  Rd = 0
)
base_states = names(state)[1:6]
dist_states = base_states %+% "d"
strats = c("", "d") # distancing strategies
ramp_period = seq(from = ymd(20210315), to = ymd(20210322), by = 1)
time_ratio = rev((seq_along(ramp_period) - 1) / (length(ramp_period) - 1))
f2 = 0.22
params_timevar = data.frame(
  Date = ramp_period,
  Symbol = rep("f", length(ramp_period)),
  Value = f2 + time_ratio * (1 - f2),
  Type = rep("abs", length(ramp_period))
)
model = (
  flexmodel(
    params = params,
    state = state,
    start_date = "2021-02-01",
    end_date = "2021-06-01",
    do_hazard = TRUE,
    params_timevar = params_timevar
  )
  
  # flow between distancing strategies
  %>% rep_rate(base_states, dist_states, ~ (ud))
  %>% rep_rate(dist_states, base_states, ~ (ur))
  
  # force of infection
  %>% vec_rate(
    "S"  %+% strats, 
    "E1" %+% strats, 
    foi
  )
  
  # flow within distancing strategies
  %>% rep_rate(
    "E1" %+% strats,
    "E2" %+% strats,
    ~ (k1)
  )
  %>% rep_rate(
    "E2" %+% strats,
    "I"  %+% strats,
    ~ (k2)
  )
  %>% rep_rate(
    "I" %+% strats,
    "Q" %+% strats,
    ~ (q)
  )
  %>% rep_rate(
    expand_names(c("I", "Q"), strats, ""),
    expand_names(c("R", "R"), strats, ""),
    ~ (1/D)
  )
)
(model
  %>% simulation_history
  %>% select(Date, matches(any_var(model$state)))
  %>% pivot_longer(!Date)
  %>% ggplot()
   +  facet_wrap(~ name, scales = 'free', ncol = 2)
   +  geom_line(aes(Date, value))
)
```

## BC Covid Omicron

The BC group also developed a model with two strains for the Omicron wave of Covid-19.

*DISCLAIMER: This is not meant to illustrate a realistic forecast, but rather to illustrate how the basic model structure can be expressed in McMasterPandemic*

```{r make_init_omicron, echo=FALSE}

make_init = function(params,
                     N=5.07e6, vaxlevel = 0.8,
                      port_wane = 0.04,
                      past_infection = 0.1,
                      incres = 300,
                      incmut = 10) {
  params_list = as.list(params)
  ff=6/7 # fudge factor . hard to get incidence right since it depends on other parameters too (2/3)
  Vtot = vaxlevel*N*(1-port_wane) # allocate to V, Ev, Iv
  Wtot = vaxlevel*N*port_wane # allocate to W, Ew, Iw
  # some have had covid. but they might also have been vaccinated.
  # set up Rs
  R0 = N*past_infection*(1-vaxlevel)  # past infections, but not vaccinated
  Rv0 = N*past_infection*(vaxlevel) * (1-port_wane) # recovered, vaxd and not waned
  Rw0 = N*past_infection*(vaxlevel) * port_wane # rec, vaxd, waned
  # set up Es : resident
  Ertot = ff*incres/params_list$sigma # total Er
  Ervw = vaxlevel*params_list$ve*Ertot # vaccinated
  Erw0 = Ervw * port_wane # waned
  Erv0 = Ervw *(1-port_wane) # not waned.  these two add to Ervwboth
  Er0 = (1-vaxlevel*params_list$epsilon_r)*Ertot # unvaccinated
  # set up Es : mutant
  Emtot = ff*incmut/params_list$sigma # total Er
  Emvw = vaxlevel*params_list$ve*Emtot # vaccinated
  Emw0 = Emvw * port_wane # waned
  Emv0 = Emvw *(1-port_wane) # not waned.  these two add to Ervwboth
  Em0 = (1-vaxlevel*params_list$epsilon_m)*Emtot # unvaccinated
  # set up Is. for now just make them 2x Es since they last twice as long
  Ir0=ff*2*Er0; Irv0 = ff*2*Erv0; Irw0 = ff*2*Erw0
  Im0 = ff*2*Em0; Imv0=ff*2*Emv0; Imw0= ff*2*Emw0

  # set up V0, W0
  # first line plus Rv0 adds to the V total but we have to leave room for the E, I
  V0 = N*(1-past_infection)*vaxlevel*(1-port_wane) -
    (Erv0+Irv0+Emv0+Imv0)*(1-port_wane)
  W0 = N*(1-past_infection)*vaxlevel*(port_wane) -
    ( Erw0+Irw0+Emw0+Imw0)*(port_wane)
  # set up S
  initmost = c(Er=Er0, Em =Em0, Ir=Ir0, Im=Im0, R=R0,
               V=V0, Erv=Erv0, Emv=Emv0, Irv = Irv0,  Imv=Imv0, Rv=Rv0,
               W=W0, Erw=Erw0, Emw=Emw0, Irw = Irw0,  Imw=Imw0, Rw=Rw0)
  state = c(S=N-sum(initmost), initmost)
  return(state)
}
```
```{r omicron, fig.height=8, fig.width=8}
params = c(
  sigma=1/3, # incubation period (3 days) (to fixed)
  gamma=1/(5), #recovery rate (fixed)
  nu =0.007, #vax rate: 0.7% per day (fixed)
  mu=1/(82*365), # 1/life expectancy (fixed)
  w1 = 1/(3*365),# waning rate from R to S (fixed)
  w2 = 1/(3*365), # waning rate from Rv to V (fixed)
  w3 = 1/(3*365),# waning rate Rw to W (fixed)
  ve=1, # I think this should be 1. it is not really efficacy  ( fixed)
  beta_r=0.72, #transmission rate (to estimate) (0.35)
  beta_m=0.8*2.2, #transmission rate (to estimate)(*1.9)
  epsilon_r = (1-0.8), # % this should be 1-ve
  epsilon_m = (1-0.6), # % escape capacity #(fixed)
  b= 0.006, # booster rate  (fixed)
  beff = 0.7, # booster efficacy
  wf=0.2, # protection for newly recoverd #0.2
  N=5e6,
  E0=5,
  S0=1-1e-5,
  c=1
)

# dimensions of model structure
vax_states = c("unvax", "onedose", "boost")
variant_states = c("delta", "omicron")
epi_states = c("S", "E" %_% variant_states, "I" %_% variant_states, "R")

# vectors representing variant model structure
I_delta   = vec("I" %_% "delta"   %_% vax_states)
I_omicron = vec("I" %_% "omicron" %_% vax_states)
E_delta   = vec("E" %_% "delta"   %_% vax_states)
E_omicron = vec("E" %_% "omicron" %_% vax_states)

# initial state vector
state = layered_zero_state(epi_states, vax_states)
all_states = names(state)
state[] = unname(make_init(params))
print(state)

two_strain_bc = (flexmodel(
    params = params,
    state = state,
    start_date = "2022-01-01",
    end_date = "2022-05-01",
    do_hazard = TRUE
  )

  # sum over every vax status to get
  # total numbers in I boxes for each
  # variant
  %>% add_state_param_sum("I_delta", "I_delta_")
  %>% add_state_param_sum("I_omicron", "I_omicron_")

  # R to S boxes for every vax status
  #   -- waning
  %>% vec_rate(
    from = "R" %_% vax_states,
    to   = "S" %_% vax_states,
    vec('w1', 'w2', 'w3')
  )

  # S_delta to E_delta for every vax status
  #   -- delta force of infection
  %>% vec_rate(
    from = "S"             %_% vax_states,
    to   = "E" %_% "delta" %_% vax_states,
    struc("(c) * (beta_r) * (1/N) * (I_delta)") * vec("1", "epsilon_r", "epsilon_r")
  )

  # S_omicron to E_omicron for every vax status
  #   -- omicron force of infection
  %>% vec_rate(
    from = "S"               %_% vax_states,
    to   = "E" %_% "omicron" %_% vax_states,
    struc("(c) * (beta_m) * (1/N) * (I_omicron)") * vec("1", "epsilon_m", "epsilon_m")
  )

  # R to E_delta for every vax status
  #   -- delta force of infection of recovered individuals
  %>% rep_rate(
    from = "R"             %_% vax_states,
    to   = "E" %_% "delta" %_% vax_states,
    ~ (wf) * (epsilon_r) * (c) * (beta_r) * (1/N) * (I_delta)
  )

  # R to E_omicron for every vax status
  #   -- omicron force of infection of recovered individuals
  %>% rep_rate(
    from = "R"               %_% vax_states,
    to   = "E" %_% "omicron" %_% vax_states,
    ~ (wf) * (epsilon_m) * (c) * (beta_m) * (1/N) * (I_omicron)
  )

  # E to I for all variant-vax combinations
  %>% rep_rate(
    from = "E" %_% expand_names(variant_states, vax_states),
    to   = "I" %_% expand_names(variant_states, vax_states),
    ~ (sigma)
  )

  # recovery for all variant-vax combinations
  %>% rep_rate(
    from = "I" %_% expand_names(variant_states, vax_states),
    to   = "R" %_% rep(vax_states, each = length(variant_states)),
    ~ (gamma)
  )

  # demographics
  %>% rep_rate(
    from = all_states,
    to   = "S_unvax",
    ~ (mu)
  )

  # vaccination
  %>% add_rate(from = "S_unvax",   to = "S_onedose", ~ (nu) * (ve))
  %>% add_rate(from = "S_onedose", to = "S_boost",   ~ (b)  * (ve))
)

category_pattern = "^(S|E|I|R)(_omicron|_delta)?_(unvax|onedose|boost)$"
(two_strain_bc
  %>% simulation_history
  %>% select(Date, matches(category_pattern))
  %>% pivot_longer(-Date, names_to = "Compartment", values_to = "State")
  %>% mutate(`Epi Status` = sub(category_pattern, '\\1\\2', Compartment, perl = TRUE))
  %>% mutate(`Vaccination Status` = sub(category_pattern, '\\3', Compartment, perl = TRUE))
  %>% arrange(`Epi Status`, `Vaccination Status`)
  %>% mutate(Compartment = factor(Compartment, all_states))
  %>% ggplot()
   +  facet_wrap(
     ~ Compartment, ncol = 3, scales = 'free', dir = 'v')
   +  geom_line(aes(Date, State))
)
```

## Classic McMasterPandemic

```{r classic-macpan}
params = read_params("ICU1.csv")
model = (flexmodel(
    params = params,
    state = make_state(params = params),
    start_date = "2020-03-10",
    end_date = "2020-12-10",
    do_make_state = FALSE,
    do_hazard = TRUE
  )
  %>% add_rate("E", "Ia", ~ (alpha) * (sigma))
  %>% add_rate("E", "Ip", ~ (1 - alpha) * (sigma))
  %>% add_rate("Ia", "R", ~ (gamma_a))
  %>% add_rate("Ip", "Im", ~ (mu) * (gamma_p))
  %>% add_rate("Ip", "Is", ~ (1 - mu) * (gamma_p))
  %>% add_rate("Im", "R", ~ (gamma_m))
  %>% add_rate("Is", "H", ~
                 (1 - nonhosp_mort) * (phi1) * (gamma_s))
  %>% add_rate("Is", "ICUs", ~
                 (1 - nonhosp_mort) * (1 - phi1) * (1 - phi2) * (gamma_s))
  %>% add_rate("Is", "ICUd", ~
                 (1 - nonhosp_mort) * (1 - phi1) * (phi2) * (gamma_s))
  %>% add_rate("Is", "D", ~ (nonhosp_mort) * (gamma_s))
  %>% add_rate("ICUs", "H2", ~ (psi1))
  %>% add_rate("ICUd", "D", ~ (psi2))
  %>% add_rate("H2", "R", ~ (psi3))
  %>% add_rate("H", "R", ~ (rho))
  %>% add_rate("Is", "X", ~ (1 - nonhosp_mort) * (phi1) * (gamma_s))
  %>% add_rate("S", "E", ~
      (Ia) * (beta0) * (1 / N) * (Ca) +
      (Ip) * (beta0) * (1 / N) * (Cp) +
      (Im) * (beta0) * (1 / N) * (Cm) * (1 - iso_m) +
      (Is) * (beta0) * (1 / N) * (Cs) * (1 - iso_s))
  %>% add_outflow(".+", "^(S|E|I|H|ICU|D|R)")
)
(model
  %>% simulation_history
  %>% pivot_longer(!Date)
  %>% mutate(name = factor(name, levels = topological_sort(model)))
  %>% ggplot()
   +  facet_wrap(~ name, scales = 'free', ncol = 3)
   +  geom_line(aes(Date, value))
)
```

## Granich HIV Model

The HIV model from [Granich et al (2008)](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(08)61697-9/fulltext).
```{r granich}
start_date = ymd(20220501)
end_date = start_date + days(30)

params = c(
  # phenomenological heterogeneity parameters and constants
  alpha = 0.1,
  lambda0 = 0.002,
  n = 1,  # exponential decrease when n = 1
  e = exp(1),
  minus_1 = -1,

  # background birth and mortality rates
  beta = 0.02,
  mu = 0.01,
  one = 1,

  # treatment efficacy (smaller is less infectious)
  epsilon = 0.2,

  # disease progression rate
  rho = 0.15,

  # treatment rate
  tau = 0.8, # value from paper

  # treatment stopping rate
  phi = 0.015, # value from paper

  # disease progression rate for treated individuals
  sigma = 0.1
)

number_of_stages = 4
state = c(
  S = 999,
  layered_zero_state("I", 1:number_of_stages),
  layered_zero_state("A", 1:number_of_stages),
  D_disease = 0,
  D_background = 0
  #birth_pool = 1000
)
state["I_1"] = 1

alive_state_nms = grep("^(S|I|A)", names(state), value = TRUE)
Ivec = vec("I" %_% 1:number_of_stages)
Avec = vec("A" %_% 1:number_of_stages)
epsilon = struc("epsilon")
I = as.character(sum(Ivec) + sum(Avec))
J = as.character(sum(Ivec) + sum(epsilon * Avec))

granich = (flexmodel(
  params = params,
  state = state,
  start_date = start_date,
  end_date = end_date,
  do_hazard = TRUE
)
  %>% add_state_param_sum("N", "^(S|I|A)")
  %>% add_factr("I", I)
  %>% add_factr("J", J)
  %>% add_factr("minus_alpha", ~ (alpha) * (minus_1))
  %>% add_factr("P", ~ (I) * (1/N))
  %>% add_pow("het_exponent", "P", "n", "minus_alpha")
  %>% add_pow("lambda", "e", "het_exponent", "lambda0")

  %>% rep_rate(
    from = alive_state_nms,
    to   = "S",
    ~ (beta)
  )

  %>% rep_rate(
    from = alive_state_nms,
    to = "D_background",
    ~ (mu)
  )

  # transmission
  %>% add_rate(
    from = "S",
    to = "I_1",
    ~ (lambda) * (S) * (J) * (1/N)
  )

  # treatment dynamics
  %>% rep_rate(
    from = "I" %_% (1:number_of_stages),
    to   = "A" %_% (1:number_of_stages),
    ~ (tau)
  )
  %>% rep_rate(
    from = "A" %_% (1:number_of_stages),
    to   = "I" %_% (1:number_of_stages),
    ~ (phi)
  )

  # disease progression
  %>% rep_rate(
    from = "I" %_% (1:(number_of_stages - 1)),
    to   = "I" %_% (2:number_of_stages),
    ~ (rho)
  )
  %>% rep_rate(
    from = "A" %_% (1:(number_of_stages - 1)),
    to   = "A" %_% (2:number_of_stages),
    ~ (sigma)
  )

  # disease death
  %>% add_rate("I" %_% number_of_stages, "D_disease", ~ (rho))
  %>% add_rate("A" %_% number_of_stages, "D_disease", ~ (sigma))

  #%>% add_outflow(from = "^[A-Z]") # don't flow out of the 'birth pool'
  %>% add_outflow(
    from = '.+',
    to = '^(I|A|D)_'
  )

  %>% add_sim_report_expr("Itotal", sum(Ivec))
  %>% add_sim_report_expr("Atotal", sum(Avec))
)

(granich
  %>% simulation_history
  %>% select(Date, S, Itotal, Atotal, J, D_background, D_disease, lambda, N, P)
  %>% pivot_longer(-Date)
  %>% ggplot
  + facet_wrap(~ name, ncol = 3, scales = 'free')
  + geom_line(aes(Date, value))
)
```
